Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (\emph{Sida fallax})

### 2023-09-07 ms/ms.tex
final checks before submission

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden, enhanced]}{\end{tcolorbox}}\fi

@@

\textbf{Methods} We developed a new method to quantify ``amphistomy
advantage'', \(\mathrm{AA}\), as the ratio of photosynthesis in an
amphistomatous leaf to that of the same leaf with the same total
stomatal conductance but with gas exchange blocked through the upper
(adaxial) surface, which we term ʻpseudohypostomyʻ. We thus estimated
\(\mathrm{AA}\) and related physiological and anatomical traits in 12
populations, six coastal (open, sunny) and six montane (closed, shaded),
of the indigenous Hawaiian species ʻilima (\emph{Sida fallax}).

\textbf{Key results} Coastal ʻilima leaves benefit 4.58 times more from
amphistomy compared to their montane counterparts. Our evidence was
equivocal with respect to two hypotheses -- that coastal leaves benefit
more because they are thicker and therefore have lower CO\(_2\)
conductance through the internal airspace, and that leaves with more
similar conductances at each surface benefit more than leaves with few
stomata on the upper surface.

@@

Amphistomy, the presence of stomata on both lower and upper surfaces of
broad leaves, should increase carbon gain by reducing the average
diffusion pathlength between stomata and chloroplasts
(\autoref{fig:compare-circuit-diagrams}), yet paradoxically this
seemingly simple adaptation is uncommon in nature and we don't know why.
Understanding variation in stomatal traits like amphistomy is imperative
because these tiny pores play an outsized ecological role in the global
carbon and water cycles (Hetherington and Woodward, 2003; Berry et al.,
2010). A widely applicable, accurate representation of how stomata
mediate the relationship between CO\(_2\) gained through photosynthesis
and water lost through evapotranspiration is essential to predict future
climate using Earth Systems Models (Jarvis, 1976; Ball et al., 1987;
Collatz et al., 1991; Leuning, 1995; Sellers et al., 1997). Optimality
models accurately predict the major cause of water loss, stomatal
conductance (\(g_\mathrm{sw}\)), by assuming plants maximize carbon gain
minus a cost of water (Cowan and Farquhar, 1977; Givnish, 1986; Medlyn
et al., 2011; Lin et al., 2015; Wang et al., 2017; Franks et al., 2018;
Deans et al., 2020; Franklin et al., 2020; Wang et al., 2020; Harrison
et al., 2021). Despite the success of optimality modeling in predicting
\(g_\mathrm{sw}\), the same modeling approach has so far failed to
explain the rarity of amphistomatous leaves (Muir, 2019). \textbf{This
gap between theory and observations strongly implies that we remain
ignorant about some key benefits and costs associated with stomata.}

@@

The most promising adaptive hypothesis is that amphistomy is important
for maximizing photosynthetic rate under high light. Mott et al. (1982)
proposed that ``plants with a high photosynthetic capacity, living in
full-sun environments, and experiencing rapidly fluctuating or
continuously available soil water'' would benefit most, in terms of
increased carbon gain, from having amphistomatous leaves. As described
above, herbs from sunny habitats are often amphistomatous. Most
variation in stomatal density ratio (\(\mathrm{SR}\), the ratio of
stomatal density between the upper and lower surfaces) among species is
assumed to be genetic, but there is also putatively adaptive plasticity
in response to light. Leaves of \emph{Ambrosia cordifolia}, a desert
perennial herb, are hypostomatous under low light (photosynthetic photon
flux density, PPFD = 110 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\))
but develop \textasciitilde20\% of their stomata on the upper surface
under high light (1700 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\))
(Mott and Michaelson, 1991). Similarly, \emph{Solanum lycopersicum}
leaves are hypostomatous when grown in the shade but develop
\textasciitilde20\% of their stomata on the upper surface grown under
high light-intensity (Gay and Hurd, 1975). Adult leaves of
\emph{Eucalyptus globulus} are amphistomatous, but the proportion of
adaxial stomata increases from \textasciitilde10-20\% under low light to
\textasciitilde30-40\% under high light (James and Bell, 2001). In
summary, both genetic and plastic responses to evince a widespread
association between light and \(\mathrm{SR}\).

The association between high light and amphistomy suggests that `sun'
leaves have the most to gain in terms of increased photosynthesis from
having stomata on both surfaces, as Mott et al. (1982) hypothesized.
Parkhurst (1978) proposed quantifying this benefit as `amphistomy
advantage' (\(\mathrm{AA}\)), which we adopt here with some modification
(see Materials and Methods). This hypothesis has never been tested
directly by comparing the photosynthetic rate of an amphistomatous leaf
to that of an otherwise identical hypostomatous leaf with the same total
stomatal conductance under the same conditions. We propose a
straightforward method to do this by experimentally creating a
pseudohypostomatous leaf with gas exchange blocked through the upper
surface (see Materials and Methods). One reason that sun leaves might
have greater \(\mathrm{AA}\) is that they are usually thicker or denser
(Poorter et al., 2019), resulting in lower \(g_\mathrm{ias}\)
(Parkhurst, 1978). A nonmutually exclusive hypothesis is that if sun
leaves have a stomatal density ratio closer to 0.5 (same density on each
leaf surface), this will confer a greater advantage than an
amphistomatous leaf with most stomata on one surface. In other words,
amphistomy doesn't make much difference if one leaf surface has very
open stomata on it. We therefore predict that sun leaves will have
greater \(\mathrm{AA}\) possibly because they have thicker leaves and/or
\(\mathrm{SR}\) closer to 0.5. We actually report
\(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
that takes into account differences in both stomatal density and size
between surfaces.

The native flora of the Hawaiian archipelago is a excellent system to
test the relationship between light habitat and \(\mathrm{AA}\). Many
lineages have adapted to different light habitats after colonization and
leaf anatomical traits such as \(\mathrm{SR}\) and thickness vary within
and among closely related species. It is hypothesized that the common
ancestor in many Hawaiian clades was a ``weedy'' species with high
dispersal ability adapted to open habitats (Carlquist, 1966).
Colonization was followed by adaptive radiation into higher elevation,
montane, closed, forested habitats. Consequently, adaptation to sun and
shade is a common axis of phenotypic variation among Hawaiian plants
such as lobeliads (Givnish et al., 2004; Montgomery and Givnish, 2008;
Givnish et al., 2009; Givnish and Montgomery, 2014; Scoffoni et al.,
2015), \emph{Bidens} (Carlquist, 1966; Knope et al., 2020),
\emph{Scaevola} (Robichaux and Pearcy, 1984; McKown et al., 2016),
\emph{Euphorbia} (Sporck, 2011), and \emph{Plantago} (Dunbar-Co et al.,
2009).

Here we focus on variation within an indigenous plant species \emph{Sida
fallax} Walp. (Malvaceae), known in the Hawaiian language as ʻilima.
ʻIlima is found from sea level to elevations \(>1000\) mas on multiple
Hawaiian islands. Coastal populations are morphologically different from
montane populations (\autoref{fig:study-system}). Coastal regions of
Hawaiʻi are characterized by high sun exposure, warmer temperatures,
high winds, salinity, and variation in water availability. Coastal
populations of ʻilima tend to be short and prostrate which likely helps
them to withstand the windy environment (\autoref{fig:study-system}a).
The leaves of these populations are covered on both surfaces in dense,
soft hairs that give the leaves a silvery green appearance
(\autoref{fig:study-system}b), which helps mitigate water loss by
reflecting solar radiation, thereby lowering leaf temperature
(Ehleringer and Björkman, 1978). Montane regions, on the other hand,
provide very different challenges. Many other tall species grow on the
slopes of these wet mountainous regions, which makes light competition a
factor that plants may need to adapt to. Possibly due to this, montane
populations are erect and shrub- or tree-like, capable of growing meters
tall with strong, woody stems. These individuals have smooth, green
foliage with serrated edges. Montane populations exhibit traits that
help them to compete for light availability (Raunch et al.~1997). This
montane morphology is not found in \emph{S. fallax} populations on other
Pacific Islands (Pejhanmehr, 2022). The variable morphology between
coastal and montane populations has even led some scientists to argue
that it makes up a species complex, rather than a single species (Raunch
et al.~1997).

The objectives of our study are to test whether 1) `sun' leaves of
coastal ʻilima plants will have greater \(\mathrm{AA}\) than `shade'
leaves of montane plants; and if so, is this because 2a) coastal plants
have thicker leaves than montane plants and/or 2b) coastal plants have a
\(g_{\mathrm{smax,ratio}}\) closer to 0.5?

@@

We identified 8 suitable natural populations of ʻilima on Oʻahu and 5 on
Hawaiʻi Island by consulting (Yorkston and Daehler, 2006) and citizen
scientist records on iNaturalist (Anon, 2022)
(\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites
that appeared to be cultivated. We visited sites between August and
November 2022. For logistical reasons, the sites on Hawaiʻi were sampled
in one three-day trip. We haphazardly sampled eight plants distributed
evenly between the highest and lowest elevation plants along a transect
at each site. For safety and conservation reasons, transects were along
a trail or road. We did not sample small individuals if there was risk
removing leaves would cause mortality. From each plant, we collected two
fully expanded leaves for traits measurements. We sampled stomatal
traits on all leaves; leaf thickness on one leaf from three randomly
selected plants per site; and, due to limited time, a single leaf from a
single plant at the middle of each transect for gas exchange
measurements.

@@

At each site, we selected one representative leaf from one plant near
the middle of the transect for gas exchange measurements using a
portable infrared gas analyze LI-6800PF, LI-COR Biosciences, Lincoln,
NE, USA). We estimated the photosynthetic rate (\(A\)) and stomatal
conductance to water vapor (\(g_\text{sw}\)) at saturating light
(\(\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
ambient CO\(_2\) (415 ppm), and
\(T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}\). We also estimated
`amphistomy advantage' (\(\mathrm{AA}\)) \emph{sensu} Parkhurst (1978),
but with modification. For each leaf, we measured the photosynthetic
rate of an untreated amphistomatous leaf (\(A_{\mathrm{amphi}}\)) over a
range of \(g_\text{sw}\) values, holding other environmental variables
constant. We refer to this as an \(A \textendash g_\text{sw}\) curve. We
compared the \(A \textendash g_\text{sw}\) curve of the untreated leaf
to the photosynthetic rate of ``pseudohypostomatous'' leaf
(\(A_\mathrm{hypo}\)), which is the same leaf but with gas exchange
through the upper surface blocked by a neutral density plastic
(propafilm). We refer to the untreated and partially blocked leaves as
``amphi'' and ``pseudohypo'', respectively. \(\mathrm{AA}\) is
calculated as the log-response ratio of \(A\) compared at the same total
\(g_\text{sw}\):

@@

To control for order effects, we alternated between starting with amphi
(untreated) or pseudohypo (treated) leaf measurements. In the field, we
acclimated the focal leaf to high light
(\(\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
high relative humidity (\(\mathrm{RH} = 70\%\)),
\(T_\text{leaf} = 25-30~° \text{C}\), depending on ambient temperature,
until \(A\) and \(g_\text{sw}\) reach their maximum. We used these data
as our estimates of maximum \(A\) and \(g_\text{sw}\). After that, we
decreased \(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal
closure without photosynthetic downregulation. We logged data until
\(g_\text{sw}\) reached its nadir. We then repeated the process with the
other treatment (amphi or pseudohypo).

@@

It is not feasible to record \(A_\mathrm{amphi}\) and
\(A_\mathrm{hypo}\) at the exact same \(g_\text{sw}\). To overcome this,
we fit \(A \textendash g_\text{sw}\) curves using a flexible generalized
additive model (GAM) to interpolate modeled \(A\) for amphi and
pseudohypo leaves at the same \(g_\text{sw}\). In 10 of 12 leaves, the
minimum \(g_\text{sw}\) of the amphi curve was smaller than the maximum
\(g_\text{sw}\) of the pseudohypo curve (i.e.~the curves overlapped for
a range of \(g_\text{sw}\) values). In those cases, we estimated
\(\mathrm{AA}\) at the \(g_\mathrm{sw}\) value in the middle of the
range of overlap. In 2 of 12 leaves, the \(A \textendash g_\text{sw}\)
curves did not quite overlap because the minimum \(g_\text{sw}\) of the
amphi curve was slightly greater than the maximum \(g_\text{sw}\) of the
pseudohypo curve. In those cases, we estimated \(\mathrm{AA}\) by
extrapolating slightly,
\(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
curves. We fit GAMs using the \emph{R} package \textbf{brms} version
2.20.1 (Bürkner, 2017) for Bayesian statistical inference with MCMC
sampling in \emph{Stan} (Stan Development Team, 2023). We used CmdStan
version 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
2023) to interface with \emph{R} version 4.3.1 (R Core Team, 2023). We
sampled from posterior distribution from 4 chains with 1000 iterations
each after 1000 warmup iterations. We estimated parameters and
confidence intervals as the median and 95\% quantile intervals of the
posterior, respectively. The key prediction is that
\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\).

@@

We tested whether leaf thickness (log-transformed) varied between
coastal and montane populations and among individuals within populations
using a Bayesian mixed-effects model with site-type as a fixed effect
and individual plant and site as random effects. We used the \emph{R}
package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit the model in
\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.32.2
and \textbf{cmdstanr} version 0.5.3 (Gabry and Češnovar, 2023). We
sampled the posterior distribution from 4 chains with 1000 iterations
each after 1000 warmup iterations. We estimated the relationship between
population average leaf thickness and \(\mathrm{AA}\) measured from a
single individual per population. We used this approach most of the
variation in leaf thickness was among sites (see Results) and the plant
selected for gas exchange measurements was not always among the plants
randomly selected for leaf thickness, precluding individual level
correlation. We propagated uncertainty about in \(\mathrm{AA}\) and leaf
thickness by integrating over the entire posterior distribution sample
for each variable. The key prediction is that the effect of leaf
thickness on \(\mathrm{AA}\) is positive.

@@

We tested whether \(g_\mathrm{smax,ratio}\) varied between coastal and
montane populations and among individuals within populations using a
Bayesian mutliresponse, mixed-effects model. The modeled response
variables are stomatal count and guard cell length on each surface.
Counts were modeled as negative binomially distributed variable from a
latent stomatal density and a parameter \(\phi\) to estimate
overdispersion in counts relative to a Poisson model. For all traits,
the explanatory variables were site-type as a fixed effect and leaf
within individual plant, individual plant, and site as random effects.
We used the \emph{R} package \textbf{brms} version 2.20.1 (Bürkner,
2017) to fit the model in \emph{Stan} (Stan Development Team, 2023) with
CmdStan version 2.32.2 and \textbf{cmdstanr} version 0.5.3 (Gabry and
Češnovar, 2023). We interpolated missing adaxial guard cell lengths from
6 out of 185 samples with zero adaxial stomata using the ``mi'' function
in \textbf{brms} package. We sampled the posterior distribution from 4
chains with 1000 iterations each after 1000 warmup iterations. From each
posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\) as:

@@

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants
more than that of montane plants. \(\mathrm{AA}\) was significantly
greater than 0 (95\% confidence intervals did not overlap 0) in 5 of 6
coastal leaves, but only 1 of 6 montane leaves
(\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual
curves). Overall, the average \(\mathrm{AA}\) among coastal and montane
leaves is \(0.12~[\numrange{0.078}{0.15}]\) and
\(0.027~[\numrange{-0.0033}{0.057}]\), respectively; the difference in
average \(\mathrm{AA}\) between habitat types is
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.09~[\numrange{0.04}{0.14}]\).
The difference between coastal and montane leaves is lower,
\(0.054~[\numrange{-0.012}{0.11}]\), but still greater than 0 if we
remove two sites where we extrapolated slightly beyond fitted
\(A \textendash g_\mathrm{sw}\) curves. Maximum photosynthetic rate was
slightly, but not significantly higher in coastal leaves (Welch Two
Sample \emph{t}-test, \(t_{9.65} = 1.6\), \(P = 0.14\)); total stomatal
conductance was similar (Welch Two Sample \emph{t}-test,
\(t_{9.71} = -0.09\), \(P = 0.93\)) in coastal and montane leaves
(\autoref{fig:habitat-Ags}).

@@

Custom scripts are available on a GitHub repository
(https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo
with a DOI and stable URL upon publication. Raw data will be deposited
on Dryad with a DOI and stable URL upon publication.

>>>>

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi

@@

\textbf{Methods} We developed a new method to quantify ``amphistomy
advantage'', \(\mathrm{AA}\), as the ratio of photosynthesis in an
amphistomatous leaf to that of the same leaf with the same total
stomatal conductance but with gas exchange blocked through the upper
(adaxial) surface, which we term ``pseudohypostomy''. We thus estimated
\(\mathrm{AA}\) and related physiological and anatomical traits in 12
populations, six coastal (open, sunny) and six montane (closed, shaded),
of the indigenous Hawaiian species ʻilima (\emph{Sida fallax}).

\textbf{Key results} Coastal ʻilima leaves benefit 4 times more from
amphistomy compared to their montane counterparts. Our evidence was
equivocal with respect to two hypotheses -- that coastal leaves benefit
more because they are thicker and therefore have lower CO\(_2\)
conductance through the internal airspace, and that leaves with more
similar conductances at each surface benefit more than leaves with few
stomata on the upper surface.

@@

Amphistomy, the presence of stomata on both lower and upper surfaces of
broad leaves, should increase carbon gain by reducing the average
diffusion pathlength between stomata and chloroplasts, yet paradoxically
this seemingly simple adaptation is uncommon in nature and we don't know
why. Understanding variation in stomatal traits like amphistomy is
imperative because these tiny pores play an outsized ecological role in
the global carbon and water cycles (Hetherington and Woodward, 2003;
Berry et al., 2010). A widely applicable, accurate representation of how
stomata mediate the relationship between CO\(_2\) gained through
photosynthesis and water lost through evapotranspiration is essential to
predict future climate using Earth Systems Models (Jarvis, 1976; Ball et
al., 1987; Collatz et al., 1991; Leuning, 1995; Sellers et al., 1997).
Optimality models accurately predict the major cause of water loss,
stomatal conductance (\(g_\mathrm{sw}\)), by assuming plants maximize
carbon gain minus a cost of water (Cowan and Farquhar, 1977; Givnish,
1986; Medlyn et al., 2011; Lin et al., 2015; Wang et al., 2017; Franks
et al., 2018; Deans et al., 2020; Franklin et al., 2020; Wang et al.,
2020; Harrison et al., 2021). Despite the success of optimality modeling
in predicting \(g_\mathrm{sw}\), the same modeling approach has so far
failed to explain the rarity of amphistomatous leaves (Muir, 2019).
\textbf{This gap between theory and observations strongly implies that
we remain ignorant about some key benefits and costs associated with
stomata.}

@@

The most promising adaptive hypothesis is that amphistomy is important
for maximizing photosynthetic rate under high light. Mott et al. (1982)
proposed that ``plants with a high photosynthetic capacity, living in
full-sun environments, and experiencing rapidly fluctuating or
continuously available soil water'' would benefit most, in terms of
increased carbon gain, from having amphistomatous leaves. As described
above, herbs from sunny habitats are often amphistomatous. Most
variation in stomatal density ratio (\(\mathrm{SR}\), the ratio of
stomatal density between the upper and lower surfaces) among species is
assumed to be genetic, but there is also putatively adaptive plasticity
in response to light. Leaves of \emph{Ambrosia cordifolia}, a desert
perennial herb, are hypostomatous under low light (photosynthetic photon
flux density, PPFD = 110 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\))
but develop \textasciitilde20\% of their stomata on the upper surface
under high light (1700 \(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\))
(Mott and Michaelson, 1991). Similarly, \emph{Solanum lycopersicum}
leaves are hypostomatous when grown in the shade but develop
\textasciitilde20\% of their stomata on the upper surface grown under
high light-intensity (Gay and Hurd, 1975). Adult leaves of
\emph{Eucalyptus globulus} are amphistomatous, but the proportion of
adaxial stomata increases from \textasciitilde10-20\% under low light to
\textasciitilde30-40\% under high light (James and Bell, 2001). In
summary, both genetic and plastic responses evince a widespread
association between light and \(\mathrm{SR}\).

The association between high light and amphistomy suggests that `sun'
leaves have the most to gain in terms of increased photosynthesis from
having stomata on both surfaces, as Mott et al. (1982) hypothesized.
Parkhurst (1978) proposed quantifying this benefit as `amphistomy
advantage' (\(\mathrm{AA}\)), which we adopt here with some modification
(see Materials and Methods). This hypothesis has never been tested
directly by comparing the photosynthetic rate of an amphistomatous leaf
to that of an otherwise identical hypostomatous leaf with the same total
stomatal conductance under the same conditions. We propose a
straightforward method to do this by experimentally creating a
pseudohypostomatous leaf with gas exchange blocked through the upper
surface (see Materials and Methods). One reason that sun leaves might
have greater \(\mathrm{AA}\) is that they are usually thicker or denser
(Poorter et al., 2019), resulting in lower \(g_\mathrm{ias}\)
(Parkhurst, 1978). A nonmutually exclusive hypothesis is that if sun
leaves have a stomatal density ratio closer to 0.5 (same density on each
leaf surface), this will confer a greater advantage than an
amphistomatous leaf with most stomata on one surface. In other words,
amphistomy doesn't make much difference if one leaf surface has few open
stomata on it. We therefore predict that sun leaves will have greater
\(\mathrm{AA}\) possibly because they have thicker leaves and/or
\(\mathrm{SR}\) closer to 0.5. We actually report
\(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
that it accounts for differences in both stomatal density and size
between surfaces.

The native flora of the Hawaiian archipelago is a excellent system to
test the relationship between light habitat and \(\mathrm{AA}\). Many
lineages have adapted to different light habitats after colonization and
leaf anatomical traits such as \(\mathrm{SR}\) and thickness vary within
and among closely related species. It is hypothesized that the common
ancestor in many Hawaiian clades was a weedy species with high dispersal
ability adapted to open habitats (Carlquist, 1966). Colonization was
followed by adaptive radiation into higher elevation, montane, closed,
forested habitats. Consequently, adaptation to sun and shade is a common
axis of phenotypic variation among Hawaiian plants such as lobeliads
(Givnish et al., 2004; Montgomery and Givnish, 2008; Givnish et al.,
2009; Givnish and Montgomery, 2014; Scoffoni et al., 2015),
\emph{Bidens} (Carlquist, 1966; Knope et al., 2020), \emph{Scaevola}
(Robichaux and Pearcy, 1984; McKown et al., 2016), \emph{Euphorbia}
(Sporck, 2011), and \emph{Plantago} (Dunbar-Co et al., 2009).

Here we focus on variation within an indigenous plant species \emph{Sida
fallax} Walp. (Malvaceae), known in the Hawaiian language as ʻilima.
ʻIlima is found from sea level to elevations \(>1000\) mas on multiple
Hawaiian islands. Coastal populations are morphologically different from
montane populations (\autoref{fig:study-system}). Coastal regions of
Hawaiʻi are characterized by high sun exposure, warmer temperatures,
high winds, salinity, and variation in water availability. Coastal
populations of ʻilima tend to be short and prostrate which likely helps
them to withstand the windy environment (\autoref{fig:study-system}a).
The leaves of these populations are covered on both surfaces in dense,
soft hairs that give the leaves a silvery green appearance
(\autoref{fig:study-system}b), which helps mitigate water loss by
reflecting solar radiation, thereby lowering leaf temperature
(Ehleringer and Björkman, 1978). Montane regions, on the other hand,
provide very different challenges. Many other tall species grow on the
slopes of these wet mountainous regions, which makes light competition a
factor that plants may need to adapt to. Possibly due to this, montane
populations are erect and shrub- or tree-like, capable of growing meters
tall with strong, woody stems. These individuals have smooth, green
foliage with serrated edges. Montane populations exhibit traits that may
help them to compete for light availability. This montane morphology is
not found in \emph{S. fallax} populations on other Pacific Islands
(Pejhanmehr, 2022).

Because of their contrasting habitat and morphology, we treat leaves
from coastal and montane plants as representatives of sun and shade
leaves, respectively, for testing hypotheses about amphistomy advantage.
Specifically, the objectives of our study are to test whether 1) sun
leaves of coastal ʻilima plants will have greater \(\mathrm{AA}\) than
shade leaves of montane plants; and if so, is this because 2a) coastal
plants have thicker leaves than montane plants and/or 2b) coastal plants
have a \(g_{\mathrm{smax,ratio}}\) closer to 0.5?

@@

We identified 7 suitable natural populations of ʻilima on Oʻahu and 5 on
Hawaiʻi Island by consulting Yorkston and Daehler (2006) and citizen
scientist records on iNaturalist (Anon, 2022)
(\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites
that appeared to be cultivated. We visited sites between August and
November 2022. For logistical reasons, the sites on Hawaiʻi were sampled
during one three-day trip. We haphazardly sampled eight plants
distributed evenly between the highest and lowest elevation plants along
a transect at each site. For safety and conservation reasons, transects
were along a trail or road. We did not sample small individuals if there
was risk removing leaves would cause mortality. From each plant, we
collected two fully expanded leaves for trait measurements. We sampled
stomatal traits on all leaves; leaf thickness on one leaf from three
randomly selected plants per site; and, due to limited time, a single
leaf from a single plant at the middle of each transect for gas exchange
measurements.

@@

At each site, we selected one representative leaf from one plant near
the middle of the transect for gas exchange measurements using a
portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln,
NE, USA). We estimated the photosynthetic rate (\(A\)) and stomatal
conductance to water vapor (\(g_\text{sw}\)) at saturating light
(\(\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
ambient CO\(_2\) (415 ppm), and
\(T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}\). We also estimated
`amphistomy advantage' (\(\mathrm{AA}\)) \emph{sensu} Parkhurst (1978),
but with modification. For each leaf, we measured the photosynthetic
rate of an untreated amphistomatous leaf (\(A_{\mathrm{amphi}}\)) over a
range of \(g_\text{sw}\) values. We refer to this as an
\(A \textendash g_\text{sw}\) curve. We compared the
\(A \textendash g_\text{sw}\) curve of the untreated leaf to the
photosynthetic rate of pseudohypostomatous leaf (\(A_\mathrm{hypo}\)),
which is the same leaf but with gas exchange through the upper surface
blocked by a neutral density plastic (propafilm). Hypostomy refers to
leaves with stomata only present on the lower, typically abaxial,
surface. We refer to the untreated and partially blocked leaves as
``amphi'' and ``pseudohypo'', respectively. \(\mathrm{AA}\) is
calculated as the log-response ratio of \(A\) compared at the same total
\(g_\text{sw}\):

@@

To control for order effects, we alternated between starting with amphi
(untreated) or pseudohypo (treated) leaf measurements. In the field, we
acclimated the focal leaf to high light
(\(\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
high relative humidity (\(\mathrm{RH} = 70\%\)),
\(T_\text{leaf} = 25-30~° \text{C}\), depending on ambient temperature,
until \(A\) and \(g_\text{sw}\) reach their maximum. We used these data
as our estimates of maximum \(A\) and \(g_\text{sw}\). After that, we
decreased \(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal
closure without photosynthetic downregulation. All other environmental
conditions in the leaf chamber remained the same. We logged data until
\(g_\text{sw}\) reached its nadir. We then repeated the process with the
other treatment (amphi or pseudohypo).

@@

It is not feasible to record \(A_\mathrm{amphi}\) and
\(A_\mathrm{hypo}\) at the exact same \(g_\text{sw}\). To overcome this,
we fit \(A \textendash g_\text{sw}\) curves using a linear regression of
\(\text{log}(g_\mathrm{sw})\) on \(A\) to interpolate modeled \(A\) for
amphi and pseudohypo leaves at the same \(g_\text{sw}\). In 10 of 12
leaves, the minimum \(g_\text{sw}\) of the amphi curve was smaller than
the maximum \(g_\text{sw}\) of the pseudohypo curve (i.e.~the curves
overlapped for a range of \(g_\text{sw}\) values). In those cases, we
estimated \(\mathrm{AA}\) at the \(g_\mathrm{sw}\) value in the middle
of the range of overlap. In 2 of 12 leaves, the
\(A \textendash g_\text{sw}\) curves did not quite overlap because the
minimum \(g_\text{sw}\) of the amphi curve was slightly greater than the
maximum \(g_\text{sw}\) of the pseudohypo curve. In those cases, we
estimated \(\mathrm{AA}\) by extrapolating slightly,
\(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
curves. We fit Bayesian regressions using the \emph{R} package
\textbf{brms} version 2.20.1 (Bürkner, 2017) with MCMC sampling in
\emph{Stan} (Stan Development Team, 2023). We used CmdStan version
2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
the posterior distribution from 4 chains with 1000 iterations each after
1000 warmup iterations. We estimated parameters and confidence intervals
as the median and 95\% quantile intervals of the posterior,
respectively. The key prediction is that
\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\).

@@

We tested whether leaf thickness (log-transformed) varied between
coastal and montane populations and among individuals within populations
using a Bayesian mixed-effects model with habitat as a fixed effect and
individual plant and site as random effects. We used the \emph{R}
package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit the model in
\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.0
and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023). We
sampled the posterior distribution from 4 chains with 1000 iterations
each after 1000 warmup iterations. We estimated the relationship between
population average leaf thickness and \(\mathrm{AA}\) measured from a
single individual per population. We used this approach most of the
variation in leaf thickness was among sites and the plant selected for
gas exchange measurements was not always among the plants randomly
selected for leaf thickness, precluding individual level correlation. We
propagated uncertainty about in \(\mathrm{AA}\) and leaf thickness by
integrating over the entire posterior distribution sample for each
variable. The key prediction is that the effect of leaf thickness on
\(\mathrm{AA}\) is positive.

@@

We tested whether \(g_\mathrm{smax,ratio}\) varied between coastal and
montane populations and among individuals within populations using a
Bayesian mutliresponse, mixed-effects model. The modeled response
variables are stomatal count and guard cell length on each surface.
Counts were modeled as negative binomially distributed variable from a
latent stomatal density and a parameter \(\phi\) to estimate
overdispersion in counts relative to a Poisson model. For all traits,
the explanatory variables were habitat as a fixed effect and leaf within
individual plant, individual plant, and site as random effects. We used
the \emph{R} package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit
the model in \emph{Stan} (Stan Development Team, 2023) with CmdStan
version 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
2023). We interpolated missing adaxial guard cell lengths from 6 out of
185 samples with zero adaxial stomata using the ``mi'' function in
\textbf{brms} package. We sampled the posterior distribution from 4
chains with 1000 iterations each after 1000 warmup iterations. From each
posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\) as:

@@

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants
more than that of montane plants. \(\mathrm{AA}\) was significantly
greater than 0 (95\% confidence intervals did not overlap 0) in 5 of 6
coastal leaves, but only 1 of 6 montane leaves
(\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual
curves). Overall, the average \(\mathrm{AA}\) among coastal and montane
leaves is \(0.12~[\numrange{0.077}{0.15}]\) and
\(0.027~[\numrange{-0.0044}{0.057}]\), respectively; the difference in
average \(\mathrm{AA}\) between habitat types is
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.089~[\numrange{0.04}{0.14}]\).
Posterior predictions closely match observed values of \(A\)
(\autoref{fig:pp-licor}), indicating an adequate model fit from which it
we can interpolate between measurements reliably. It also suggests that
slight extrapolation beyond the data should be reliable, but this is
less certain. When we remove two leaves where extrapolated slightly
beyond fitted \(A \textendash g_\mathrm{sw}\) curves, we estimate that
\(\mathrm{AA}_\text{coastal}\) is still positive,
\(0.081~[\numrange{0.022}{0.14}]\), but the difference between coastal
and montane leaves is smaller, \(0.053~[\numrange{-0.011}{0.12}]\), and
confidence intervals slightly overlap 0. Maximum photosynthetic rate was
slightly, but not significantly higher in coastal leaves (Welch Two
Sample \emph{t}-test, \(t_{9.65} = 1.6\), \(P = 0.14\)); total stomatal
conductance was similar (Welch Two Sample \emph{t}-test,
\(t_{9.71} = -0.09\), \(P = 0.93\)) in coastal and montane leaves
(\autoref{fig:habitat-Ags}).

@@

Custom scripts are available on a GitHub repository
(https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo
with a DOI and stable URL upon publication. Raw data will be deposited
on Dryad with a DOI and stable URL upon publication. {[}THE GITHUB REPO
AND DRYAD DATA ARE AVAILABLE TO REVIEWERS{]}
